---
title: 'Calculation of total daily precipitation using ASOS data: Example'
author: "Robert McDonald"
date: "June 19, 2017"
output:
  html_document:
    keep_md: yes
  github_document: default
  pdf_document: default
---

```{r setup, include=FALSE, message=FALSE}
library(knitr)
library(XML)
library(tidyverse)
library(lubridate)
library(magrittr)
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA, 
                      results='asis', 
                      message=FALSE,
                      warning=FALSE)
```
## What is this?

The purpose of this document is to illustrate the computation of
daily precipitation totals using Automated Surface Observation
System (ASOS) [data from Iowa
State](https://mesonet.agron.iastate.edu/ASOS/).

You should be aware that the concept of a "daily precipitation
total" is fraught. If you are interested, there is a presentation on
the topic by Daryl Herzmann from Iowa State. I think the phrase
"down the rabbit hole" occurs to everyone who looks into this topic.

## History

Official historical precipitation data, by weather station, is
available from the [NOAA](https://www.ncdc.noaa.gov/cdo-web/). The
system requires you to submit a request after which they email you a
link. It is clunky but fairly quick.

```{r, echo=FALSE}
lgahist = read_csv('data/lga_daily_2013.csv')
lga = lgahist %>% 
  select(DATE, Historical=PRCP) %>% 
  mutate(year=as.numeric(substr(DATE, 1, 4)), 
         month=as.numeric(substr(DATE, 5, 6)), 
         day=as.numeric(substr(DATE, 7, 8))) %>% 
  filter(year==2013, month==6)
  

```

## Computing LGA precipitation totals from ASOS data 

The ASOS data was obtained from [Iowa
State](https://mesonet.agron.iastate.edu/request/download.phtml?network=NY_ASOS)
by selecting the New York ASOS and then LGA.

<!-- https://mesonet.agron.iastate.edu/cgi-bin/request/asos.py?station=LGA&data=all&year1=2013&month1=5&day1=30&year2=2013&month2=8&day2=1&tz=Etc%2FUTC&format=comma&latlon=no&direct=no&report_type=1&report_type=2 -->

There are two important things to realize when using the ASOS data, which is hourly.

1. The reported precipitation number is cumulative, up to a reset time which is typically between 51 and 57 minutes (thank you Daryl!) So the hourly total is this final value before the reset.

2. The daily total is the sum of the hourly values, where a day is
midnight to midnight, local *standard* time. Note that DST is not
observed for this purpose. This makes sense, as the days on which
DST is adopted and removed would otherwise create either an hour
hole or a double-counted hour.

One complication is that the reset minute can vary by site. Although
it is obvious that the LGA reset occurs at 51 minutes, this is not 
true for all ASOS. So in the code below I compute `modalminute`,
which is the most frequently occurring minute in the data. I assume 
that this is the reset time. Because R has no function for computing
the mode, I use the `tabulate` function and `which.max` to find the
most common value for minute.

Note also that I specify the time zone as "America/New_York" and
then undo DST by using the `dst` function to see if daylight savings
time is in effect, and then if so, subtracting 3600 seconds from the
time.

A final note: on two days, the offical precipitation total is off by
0.01", and this is corrected the next day. I am not sure why this
happens.

```{r}
## This was inspired by Hadley Wickham's 
## weather.r in the R package nycflights13
x = read_csv('data/asos_lga_2013-06-01_2013-08-01.csv')
x %>% 
  select(station, valid, p01i) %>% 
  mutate(date = with_tz(as.POSIXct(valid, tz='UCT'),
                       "America/New_York"),
         date = date-dst(date)*3600,
         year = year(date),
         month = month(date),
         day = day(date),
         hour = hour(date),
         minute = minute(date)) %T>% 
         {modalminute <<- which.max(tabulate(.$minute))}  %>% 
  group_by(year, month, day, hour) %>% 
  filter(minute <= modalminute) %>% 
  summarize(hourlyprecip = max(p01i)) %>% 
  group_by(year, month, day) %>% 
  summarize('Daily Precipitation (ASOS)' = sum(hourlyprecip)) %>% 
  filter(year==2013, month==6) %>% 
  left_join(lga %>% select(Historical:day)) %>% 
  kable(caption = paste0('24-hour precipitation. ',
                         'Hourly cumulation reset at ', 
                         modalminute, ' minutes'))
  
```

  